---
title: "Conducting Multiple Linear Regression in R"
author: "Adineh Yazdani"

format:
  html:
    toc: true
    number-sections: true
  pdf:
    pdf-engine: xelatex
    toc: true
    number-sections: true

bibliography: references.bib
csl: apa.csl

execute:
  echo: true
  warning: false
  message: false
---

::: center
**Hochschule Fresenius**

**Course:** Data Analysis for Decision-Making

**Professor:** Prof. Dr. Stephan Huber

**Date:** May 26, 2026
:::

------------------------------------------------------------------------

------------------------------------------------------------------------

## Scope of Presentation

This presentation covers:

- How to **load and inspect** data in R
- How to **specify and run** a multiple linear regression model using `lm()`
- How to **extract and interpret** results from `summary()`
- How to run basic **model diagnostics**

::: callout-note
Theory of regression is assumed as background knowledge.
:::

------------------------------------------------------------------------

## Workflow Overview

```{r}
#| echo: false
#| eval: true
```

::::::: columns
::: {.column width="25%"}
**1. Data Input**\
Load & inspect
:::

::: {.column width="25%"}
**2. Model Specification**\
Define formula
:::

::: {.column width="25%"}
**3. Model Estimation**\
Fit with `lm()`
:::

::: {.column width="25%"}
**4. Output Extraction**\
`summary()` & plots
:::
:::::::

> *Each step builds on the previous one — a clear, repeatable workflow.*

------------------------------------------------------------------------

## Data Input in R

Create a data frame directly in R:

```{r}
data <- data.frame(
  Salary     = c(2000, 2500, 3000, 3500, 4000),
  Age        = c(22, 25, 30, 35, 40),
  Education  = c(12, 14, 16, 16, 18),
  Experience = c(1, 3, 5, 7, 10)
)
```

::: callout-tip
In practice, import data from CSV with `read.csv("file.csv")` or from Excel with `readxl::read_excel("file.xlsx")`.
:::

------------------------------------------------------------------------

## Data Inspection

Use these built-in functions before modeling:

```{r}
head(data)      # shows the first 6 rows; confirms data loaded correctly

summary(data)   # gives min, max, mean, and quartiles for each variable

str(data)       # checks variable types (numeric, factor, etc.)

is.na(data)     # check for missing values
```

::: fragment
**Example:** `summary(data)` reveals average salary is 3000, education ranges 12–18 years, experience 1–10 years — useful context before modeling.
:::

------------------------------------------------------------------------

## Model Specification

**Formula:** `Salary ~ Age + Education + Experience`

| Symbol | Meaning |
|----|----|
| `~` | Separates outcome from predictors |
| Left side | Dependent variable **(y)**: `Salary` |
| Right side | Independent variables **(x)**: `Age`, `Education`, `Experience` |

```{r}
# The formula interface in R
Salary ~ Age + Education + Experience
```

------------------------------------------------------------------------

## Model Estimation

Fit the model using **Ordinary Least Squares (OLS)**:

```{r}
model <- lm(Salary ~ Age + Education + Experience, data = data)
```

The `lm()` call returns a **model object** containing:

- Estimated coefficients
- Residuals and fitted values
- Diagnostic information

```{r}
fitted(model)      # predicted values
residuals(model)   # observed minus predicted
summary(model)     # full results table
```

------------------------------------------------------------------------

## Extracting Results

```{r}
summary(model)
```

| Component | Interpretation |
|----|----|
| **Estimate** | Expected change in Y per one-unit increase in X, holding others constant |
| **Std. Error** | Precision of the coefficient estimate |
| **t value** | Estimate ÷ Std. Error; larger = stronger evidence |
| **Pr(\>\|t\|)** | p-value; p \< 0.05 → statistically significant |
| **R-squared** | Proportion of variance in Y explained by the model |
| **Adjusted R²** | R² penalized for number of predictors |
| **F-statistic** | Tests if at least one predictor has a non-zero coefficient |

------------------------------------------------------------------------

## Basic Diagnostics

```{r}
par(mfrow = c(2, 2))  # display 4 plots in a 2x2 grid
plot(model)
```

What `plot(model)` shows:

- **Residuals vs Fitted** — checks linearity
- **Normal Q-Q** — checks if residuals are normally distributed
- **Scale-Location** — checks constant variance (homoscedasticity)
- **Cook's Distance** — detects influential outliers

------------------------------------------------------------------------

## Implementation Example

Predict salary for a **new employee**:

```{r}
new_employee <- data.frame(
  Age        = 33,
  Education  = 16,
  Experience = 6
)

# Point estimate
predict(model, newdata = new_employee)

# With 95% prediction interval
predict(model, newdata = new_employee,
        interval = "prediction", level = 0.95)
```

::: callout-note
The prediction interval gives both a point estimate and a plausible range — useful for decision-making.
:::

------------------------------------------------------------------------

## Class Exercise {background-color="#f0f4f8"}

**Run a regression model in R using `lm()` and interpret the output.**

Use your own variables and the following structure:

```{r, eval=FALSE}
# 1. Create your dataset (at least 8 observations, 2+ predictors)
my_data <- data.frame(
  y  = c(...),   # your outcome variable
  x1 = c(...),   # first predictor
  x2 = c(...)    # second predictor
)

# 2. Fit the model
my_model <- lm(y ~ x1 + x2, data = my_data)

# 3. Inspect results
summary(my_model)

# 4. Diagnostics
par(mfrow = c(2, 2)); plot(my_model)

# 5. Predict a new observation
predict(my_model, newdata = data.frame(x1 = ..., x2 = ...))
```

**Questions to answer:** Which predictors are significant? What does adjusted R² tell you? Interpret one coefficient in plain language.

------------------------------------------------------------------------

## Key Takeaway {.center background-color="#1a3a5c"}

<br>

Multiple linear regression in R follows a **clear, repeatable workflow:**

<br>

**Load data → Inspect → Specify model with `lm()`**\
**→ Extract results with `summary()` → Check diagnostics with `plot()`**

<br>

Mastering this sequence allows you to efficiently analyze relationships between variables and interpret **coefficients**, **p-values**, and **R-squared** in your own research.

------------------------------------------------------------------------

## References

- R Core Team. (2024). *R: A language and environment for statistical computing.* R Foundation for Statistical Computing. <https://www.R-project.org/>

- James, G., Witten, D., Hastie, T., & Tibshirani, R. (2021). *An introduction to statistical learning* (2nd ed.). Springer. <https://doi.org/10.1007/978-1-0716-1418-1>

- Wooldridge, J. M. (2019). *Introductory econometrics: A modern approach* (7th ed.). Cengage Learning.

::: {.callout-note appearance="minimal"}
APA 7th edition citation style.
:::

------------------------------------------------------------------------

## Thank You {.center background-color="#1a3a5c"}

<br>

**Adineh Yazdani**\
yazdani_khomeirani.adineh\@stud.hs-fresenius.de\
Matriculation No.: 401121836

<br>

*Questions welcome!*
